```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      cache = TRUE, 
                      cache.path = "cache/3.2/")
```

# 3.2 Totals, means and ratios {-}

This section of the manual focuses on the calculation of totals, means, and ratios using data from the ESS4 survey. It begins by setting up the work environment and loading data on nonfood item expenditures. It then estimates the total and average household expenditure for each item type, along with precision measures.

Next, it calculates the share of each item's expenditure relative to total household spending, at both national and regional levels, and presents the results in a comparative table. The section concludes with a focused analysis of educational expenditure as a share of total consumption, using an aggregated dataset. All results are formatted for clear interpretation and regional comparison.


### Prepare the work environment {-}

This block resets the working environment and loads the necessary packages to continue with data processing and survey analysis in a new section. This structure ensures that each module of the analysis can be executed independently, promoting reproducibility and avoiding conflicts with previous sessions.

```{r, echo=TRUE,warning=FALSE,error=FALSE,menssage = FALSE}

#------------------------------------------------------------------------------#
#                           Cleaning R Environment                             #
#------------------------------------------------------------------------------#

rm(list = ls())
gc()

#------------------------------------------------------------------------------#
#                                Libraries                                     #
#------------------------------------------------------------------------------#

library(tidyverse)
library(survey)
library(srvyr)
library(data.table)
library(magrittr)
library(haven)
library(stringr)
library(tidyr)
library(knitr)
library(kableExtra)

# Ensure select function from dplyr is used explicitly
select <- dplyr::select

#------------------------------------------------------------------------------#
#                        Setting the RAM Memory Limit                          #
#------------------------------------------------------------------------------#

memory.limit(250000000)
```

This block loads data from section 7b of the household questionnaire, related to expenditures on nonfood items, and calculates the total expenditure per household. It then defines the survey design using the same stratification structure and primary sampling units as in previous sections, in order to apply the adjusted weights in the analysis. This step is essential to produce representative estimates of expenditure and its distribution across households.

```{r read_data, echo=TRUE,warning=FALSE,error=FALSE,menssage = FALSE}

#------------------------------------------------------------------------------#
#                           Loading Datasets                                   #
#------------------------------------------------------------------------------#

HH_data <- read_sav("data/data_ESS4/sect7b_hh_w4_v2.sav")

#------------------------------------------------------------------------------#
#                Calculating Ratio of Expenditures                            #
#------------------------------------------------------------------------------#

# Processing data
HH_data <- HH_data %>%
  mutate(expenditure = ifelse(s7q03 == 2, 0, s7q04)) %>%
  group_by(household_id) %>%
  mutate(total_expenditure = sum(expenditure, na.rm = TRUE))

# Defining the survey design
ESS4_design <- HH_data %>%
  mutate(strata = paste0(saq01, "_", saq14)) %>%
  as_survey_design(
    ids = ea_id,
    # Primary sampling unit identifier (EA)
    strata = strata,
    # Stratification by region (saq01) and urban/rural zone (saq14)
    weights = pw_w4,
    # Final adjusted weight
    nest = TRUE
  )

options(survey.lonely.psu = "fail")
summary(ESS4_design)
```

### Estimation of totals and means {-}

This block estimates the total expenditure and the average expenditure per household on nonfood items, disaggregated by item type. Using the previously defined survey design, the estimates are calculated along with their corresponding precision measures: standard error and coefficient of variation. The results are then rounded and displayed in a readable table to facilitate analysis and interpretation.

```{r resul_table, echo=TRUE, eval=FALSE, error=FALSE, message=FALSE,warning=FALSE}
#------------------------------------------------------------------------------#
# Estimation of totals and means
#------------------------------------------------------------------------------#

tab_02 <- ESS4_design %>%   
  group_by(item = item_cd_12months) %>%
  summarise(
    T_hat = survey_total(expenditure, na.rm = TRUE, vartype = c("se", "cv")),
    M_hat = survey_mean(expenditure, na.rm = TRUE, vartype = c("se", "cv"))
  )

tab_02 %>%
  transmute(
    item = as_factor(item),
    T_hat = round(T_hat),
    T_hat_cv = round(T_hat_cv * 100, 2),
    M_hat = round(M_hat),
    M_hat_cv = round(M_hat_cv * 100, 2)
  ) %>% data.frame()
```


```{r resul_table2, echo=FALSE, eval=TRUE, error=FALSE, message=FALSE,warning=FALSE}
tab_02 <- ESS4_design %>%   
  group_by(item = item_cd_12months) %>%
  summarise(
    T_hat = survey_total(expenditure, na.rm = TRUE, vartype = c("se", "cv")),
    M_hat = survey_mean(expenditure, na.rm = TRUE, vartype = c("se", "cv"))
  )

tab_02 %>%
  transmute(
    item = as_factor(item),
    T_hat = round(T_hat),
    T_hat_cv = round(T_hat_cv * 100, 2),
    M_hat = round(M_hat),
    M_hat_cv = round(M_hat_cv * 100, 2)
  ) %>% data.frame() %>% 
    kable(format = "html") %>%
  kable_styling(full_width = FALSE)
```

### Estimation of Ratio {-}

This block estimates the share of expenditure on each nonfood item relative to total household spending, both at the national level and by region, using the `survey_ratio()` function. Two tables are generated: one for Ethiopia as a whole and another disaggregated by region, both including standard errors and coefficients of variation.

```{r}
#------------------------------------------------------------------------------#
# Estimation of Ratio                                            #
#------------------------------------------------------------------------------#

tab_03_R1_total <- ESS4_design %>%
  group_by(item = item_cd_12months) %>%
  summarise(R_hat = survey_ratio(
    expenditure,
    total_expenditure,
    na.rm = TRUE,
    vartype = c("se", "cv")
  )) %>%
  mutate(Region = "ETHIOPIA")

tab_03_R1_region <-
  ESS4_design %>%
  group_by(item = item_cd_12months, Region = as_factor(saq01)) %>%
  summarise(R_hat = survey_ratio(
    expenditure,
    total_expenditure,
    na.rm = TRUE,
    vartype = c("se", "cv")
  ))

# Combining results
tab_03_R1 <- bind_rows(tab_03_R1_total, tab_03_R1_region)
```

In this block, the results are formatted as percentages and paired with their standard errors in parentheses. The table is then reshaped into wide format, displaying regions as columns to facilitate comparison.

```{r, eval=FALSE}
# Formatting output
tab_03_R1 <- tab_03_R1 %>%
  transmute(
    Region,
    item = as_factor(item),
    R_hat = round(R_hat * 100, 2),
    R_hat = paste0(R_hat, " (", round(R_hat_se * 100, 2), ")")
  ) %>%
  pivot_wider(
    names_from = Region,
    values_from = R_hat,
    values_fill = list(n = 0)
  )

tab_03_R1
```

```{r, eval=TRUE, echo=FALSE}
# Formatting output
tab_03_R1 <- tab_03_R1 %>%
  transmute(
    Region,
    item = as_factor(item),
    R_hat = round(R_hat * 100, 2),
    R_hat = paste0(R_hat, " (", round(R_hat_se * 100, 2), ")")
  ) %>%
  pivot_wider(
    names_from = Region,
    values_from = R_hat,
    values_fill = list(n = 0)
  )

tab_03_R1 %>% 
   kable(format = "html") %>%
  kable_styling(font_size = 6, full_width = FALSE)
```

### Estimating the Ratio of Educational Expenditure {-}

This block begins the analysis of educational expenditure by loading an aggregated consumption dataset. The survey design is then defined using the same stratification structure and adjusted weights applied in previous sections. This step is essential to ensure that estimates of educational spending are properly representative of the survey’s sample design.

```{r}
#------------------------------------------------------------------------------#
#                Estimating the Ratio of Educational Expenditure               #
#------------------------------------------------------------------------------#

HH_data <- read_sav("data/data_ESS4/cons_agg_w4.sav")

# Defining the survey design
ESS4_design <- HH_data %>%
  mutate(strata = paste0(saq01, "_", saq14)) %>%
  as_survey_design(
    ids = ea_id,
    # Primary sampling unit identifier (EA)
    strata = strata,
    # Stratification by region (saq01) and urban/rural zone (saq14)
    weights = pw_w4,
    # Final adjusted weight
    nest = TRUE
  )

```

#### Estimation of the ratio {-}

This block estimates the share of educational expenditure relative to total household consumption, both at the national level and by region. The survey_ratio() function is used to calculate the ratio between annual educational consumption (`educ_cons_ann`) and total annual consumption (`total_cons_ann`), including precision measures such as standard error and coefficient of variation. National and regional results are combined into a single table for further analysis and visualization.

```{r}
#------------------------------------------------------------------------------#
# Estimation of the ratio.                                                     #
#------------------------------------------------------------------------------#

tab_03_R2_total <- ESS4_design %>%
  summarise(R_hat = survey_ratio(
    educ_cons_ann,
    total_cons_ann,
    na.rm = TRUE,
    vartype = c("se", "cv")
  )) %>% mutate(Region = "ETHIOPIA")

tab_03_R2_region <- ESS4_design %>%
  group_by(Region = as_factor(saq01)) %>%
  summarise(R_hat = survey_ratio(
    educ_cons_ann,
    total_cons_ann,
    na.rm = TRUE,
    vartype = c("se", "cv")
  ))



# Combining results
tab_03_R2 <- bind_rows(tab_03_R2_region, tab_03_R2_total)

```

This final block formats the results of the educational expenditure as a share of total household consumption, expressing them as percentages. Both the estimated ratio (`R_hat`) and its standard error (`R_hat_se`) are included, allowing for easy interpretation and comparison across regions and at the national level.

```{r, echo=TRUE, eval=FALSE}
# Formatting and displaying results
tab_03_R2 %>%
  transmute(Region,
            R_hat = round(R_hat * 100, 2),
            R_hat_se = round(R_hat_se * 100, 2))  

```


```{r, echo=FALSE, eval=TRUE}
# Formatting and displaying results
tab_03_R2 %>%
  transmute(Region,
            R_hat = round(R_hat * 100, 2),
            R_hat_se = round(R_hat_se * 100, 2)) %>% 
  kable(format = "html") %>%
  kable_styling(full_width = FALSE)

```

