```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      cache = TRUE, 
                      cache.path = "cache/3.3/")
```

# 3.3 Correlations {-}

This section estimates **correlations between educational expenditure and other types of household spending** using the complex survey design from the **ESS4** dataset. It explores the relationship between educational spending and two key components of consumption: **utilities** and **food away from home**.

The estimates are presented at three levels:  
1. **National**, capturing the overall association.  
2. **By area type (urban/rural)**, to explore territorial differences.  
3. **By region**, providing a detailed subnational analysis.

The results include correlation coefficients along with their standard errors, allowing for the assessment of both the strength and precision of the associations observed.


### Prepare the work environment {-}

This block sets up the working environment, loads an aggregated individual-level consumption dataset, and defines the **survey design** for analysis. The data is stratified by region and area type (urban/rural), and both the primary sampling units (`ea_id`) and adjusted weights (`pw_w4`) are specified. Applying this structure ensures that all estimates reflect the representativeness of the survey design. Finally, a summary of the design is displayed as a preliminary validation step.

```{r, echo=TRUE,warning=FALSE,error=FALSE,menssage = FALSE}

#------------------------------------------------------------------------------#
#                           Cleaning R Environment                             #
#------------------------------------------------------------------------------#

rm(list = ls())
gc()

#------------------------------------------------------------------------------#
#                                Libraries                                     #
#------------------------------------------------------------------------------#

library(tidyverse)
library(survey)
library(srvyr)
library(data.table)
library(magrittr)
library(haven)
library(stringr)
library(tidyr)
library(knitr)
library(kableExtra)

# Ensure select function from dplyr is used explicitly
select <- dplyr::select

#------------------------------------------------------------------------------#
#                           Loading Datasets                                   #
#------------------------------------------------------------------------------#

IND_data <- read_sav("data/data_ESS4/cons_agg_w4.sav") %>%
  mutate(strata = paste0(saq01, "_", saq14))

#------------------------------------------------------------------------------#
#                       Defining Survey Design                                #
#------------------------------------------------------------------------------#
ESS4_design <- IND_data %>%
  as_survey_design(
    ids = ea_id,
    # Primary sampling unit identifier (EA)
    strata = strata,
    # Stratification by region (saq01) and urban/rural zone (saq14)
    weights = pw_w4,
    # Final adjusted weight
    nest = TRUE
  )

options(survey.lonely.psu = "fail")
summary(ESS4_design)
```

###  Estimating Correlations  {-}

This block calculates the overall correlation between annual educational expenditure and two types of spending: utilities and food away from home. The estimates account for the complex survey design.

```{r, echo=TRUE, eval=FALSE}
#------------------------------------------------------------------------------#
#                   Estimating Correlations                                    #
#------------------------------------------------------------------------------#

# Overall correlation
ESS4_design %>% summarise(
  corr1 = survey_corr(educ_cons_ann, utilities_cons_ann),
  corr2 = survey_corr(educ_cons_ann, fafh_cons_ann)
)
```
The table shows the **national-level estimated correlations** between annual educational expenditure and two types of spending:  

- **`corr1`**: correlation between educational spending and utility expenses, estimated at **0.19**, indicating a **moderate positive association**.  
- **`corr2`**: correlation between educational spending and food away from home, estimated at **0.06**, suggesting a **weak association**.  

Both estimates include their respective **standard errors**, allowing for an assessment of precision within the context of the complex survey design.


```{r, echo=FALSE, eval=TRUE}
ESS4_design %>% summarise(
  corr1 = survey_corr(educ_cons_ann, utilities_cons_ann),
  corr2 = survey_corr(educ_cons_ann, fafh_cons_ann)
) %>%  kable(format = "html") %>%
  kable_styling(full_width = FALSE)
```

This block estimates the correlation by area type (urban or rural), allowing us to identify differences in spending patterns depending on geographic setting.

```{r, echo=TRUE, eval=FALSE}
# Correlation grouped by urban/rural area
ESS4_design %>%
  group_by(area = as_factor(saq14)) %>%
  summarise(
    corr1 = survey_corr(educ_cons_ann, utilities_cons_ann),
    corr2 = survey_corr(educ_cons_ann, fafh_cons_ann)
  ) 
```

The table shows the **correlations between educational expenditure and other types of spending**, broken down by area of residence:

- In **rural areas**, the correlation between educational spending and utilities (`corr1`) is **moderate (0.19)** but has **higher uncertainty** (standard error = 0.073). The correlation with food away from home (`corr2`) is **nearly zero (0.009)**.
  
- In **urban areas**, the correlation with utilities remains **positive (0.16)** and is more precise (standard error = 0.025). The correlation with food away from home is **slightly higher (0.039)** than in rural areas but still weak.

These results suggest that the link between educational and utility spending is consistent across areas, while the relationship with food away from home is weak—particularly in rural settings.


```{r, echo=FALSE, eval=TRUE}
ESS4_design %>%
  group_by(area = as_factor(saq14)) %>%
  summarise(
    corr1 = survey_corr(educ_cons_ann, utilities_cons_ann),
    corr2 = survey_corr(educ_cons_ann, fafh_cons_ann)
  ) %>%  kable(format = "html") %>%
  kable_styling(full_width = FALSE)
```

This block estimates the correlation between expenditure variables by region, providing a more detailed view of consumption relationships at the subnational level.

```{r, echo=TRUE, eval=FALSE}
# Correlation grouped by region
ESS4_design %>% 
  group_by(Region = as_factor(saq01)) %>%
  summarise(
    corr1 = survey_corr(educ_cons_ann, utilities_cons_ann), 
    corr2 = survey_corr(educ_cons_ann, fafh_cons_ann)
  )

```

The table displays **regional correlations** between educational expenditure and two types of spending: **utilities (corr1)** and **food away from home (corr2)**, along with their standard errors.

- In most regions, the correlation between **education and utilities** is **positive and moderate**, with the highest values in **Dire Dawa (0.33)**, **Somali (0.31)**, **Amhara (0.24)**, and **Harar (0.23)**. This suggests that households investing more in education also tend to spend more on utilities in those areas.

- The correlation with **food away from home** is **weaker and more variable**. Notably, **Tigray (0.25)** shows a stronger positive association, while **Benishangul Gumuz (-0.07)** and **Dire Dawa (-0.04)** show **negative correlations**, possibly reflecting different spending trade-offs or priorities.

- The **standard errors** indicate precision: **Addis Ababa** has the most precise estimates, while regions like **SNNP** and **Tigray** show greater uncertainty.


```{r, echo=FALSE, eval=TRUE}
# Correlation grouped by region
ESS4_design %>% 
  group_by(Region = as_factor(saq01)) %>%
  summarise(
    corr1 = survey_corr(educ_cons_ann, utilities_cons_ann), 
    corr2 = survey_corr(educ_cons_ann, fafh_cons_ann)
  ) %>%  kable(format = "html") %>%
  kable_styling(full_width = FALSE)

```
