---
title: "Appendix of Chapter 9 - Analysis of Household Surveys - Examples in R"
author: ""
date: "`r Sys.Date()`"
documentclass: book
bibliography: chapterbib.bib
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
fontsize: 12pt
geometry: margin=3cm
github-repo: psirusteam/chapter9Rcodes
description: "CHAPTER 9 - EXAMPLES IN R"
knit: "bookdown::render_book"
linkcolor: blue
output: 
  bookdown::pdf_book:
    toc: true
    toc_depth: 1
    keep_tex: true
    latex_engine: xelatex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      cache = TRUE)

library("dplyr")
library("tidyverse")
library("readstata13") 
library("survey")
library("srvyr")
library("ggplot2")
library("TeachingSampling")
library("samplesize4surveys")
library(convey)
library(rstan)
library(printr)
library(knitr)
rstan_options (javascript = FALSE)
library(thematic)
library(ggplot2) 
theme_set(theme_bw())
thematic_on(
  bg = "white", fg = "black", accent = "red",
  font = font_spec("Oxanium", scale = 1.25)
)

## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
## bookdown::render_book("index.Rmd", "bookdown::word_document2")
```

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      cache = TRUE, 
                      cache.path = "cache/index/")
```

# Abstract {-}

This repository hosts the appendix for Chapter 9 (*Analysis of Household Surveys*) of the *United Nations Handbook on Surveys of Individuals and Households*. This appendix enables readers to reproduce the examples and results presented in the main document by following the computational steps outlined here.   

Chapter 9 focuses on the analysis of survey data. To provide practical examples, we utilize real household survey data from the World Bank's *Living Standards Measurement Study (LSMS)* project. Each section of Chapter 9 is accompanied by corresponding examples, which illustrate and expand upon the key concepts discussed in the chapter.  

This appendix was developed by the Statistics Division of the United Nations Economic Commission for Latin America and the Caribbean (UN-ECLAC). The implementation was commissioned to Andrés Gutiérrez (<andres.gutierrez@eclac.org>), Technical Adviser at UN-ECLAC Statistics Division, and Stalyn Guerrero (<guerrerostalyn@gmail.com>), Consultant of the same Division.  











